---
title: "Simple Linear Regression: CPU Cores vs. Performance"
output: ioslides_presentation
---

<style type="text/css">
body p, div, h1, h2, h3, h4, h5 {
 color: black;
 font-family: Modern Computer Roman;
}
slides > slide.title-slide hgroup h1 {
 color: #8C1D40;
}
h2 {
 color: #8C1D40;
}
</style>

## Data Preparation
```{r setup, include=FALSE}
library(ggplot2)
library(plotly)
library(knitr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

cpu_data <- read.csv("CPU.csv") 
clean_data <- na.omit(cpu_data[, c("cores", "cpuMark")])
```

Dataset: `r nrow(cpu_data)` CPUs

Variables: cores (X) and cpuMark (Y)

## Simple Linear Regression Model

$$Y = \beta_0 + \beta_1 X + \epsilon$$

- $Y$ = cpuMark
- $X$ = cores
- $\beta_0$ = intercept
- $\beta_1$ = slope

## Scatter Plot
```{r plot1, echo=FALSE, fig.height=4.5}
ggplot(clean_data, aes(x = cores, y = cpuMark)) +
  geom_point(alpha = 0.5, color = "#8C1D40") +
  geom_smooth(method = "lm", color = "#FFC627") +
  theme_minimal()
```

## Fitting the Model
```{r model, echo=TRUE}
model <- lm(cpuMark ~ cores, data = clean_data)
summary(model)
```

## Model Equation
```{r coef, echo=FALSE}
b0 <- round(coef(model)[1], 2)
b1 <- round(coef(model)[2], 2)
```

$$\hat{Y} = `r b0` + `r b1` X$$

R² = `r round(summary(model)$r.squared, 3)`

## Residual Plot
```{r plot2, echo=FALSE, fig.height=4.5}
ggplot(data.frame(fitted = fitted(model), resid = residuals(model)), 
       aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5, color = "#8C1D40") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal()
```

## 3D Visualization
```{r plotly, echo=FALSE, fig.height=4.5}
plot_data <- na.omit(cpu_data[, c("cores", "cpuMark", "price")])
plot_data <- plot_data[plot_data$price > 0, ]

plot_ly(plot_data, x = ~cores, y = ~price, z = ~cpuMark,
        type = "scatter3d", mode = "markers",
        marker = list(size = 3, color = ~cpuMark, colorscale = "Viridis"))
```

## Conclusion

More cores = higher performance

Slope: `r b1` points per core

Model fits well (R² = `r round(summary(model)$r.squared, 3)`)