---
title: "Project 1: NVIDIA (NVDA) Stock Analysis"
author: "Jooho Kim"
date: "2025-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Background and Problem Definition

NVIDIA (NVDA), a central player in the AI boom, has seen dynamic but highly volatile stock price growth, especially post-2020. This volatility makes forecasting challenging, as the price is driven not only by historical patterns but also by external market sentiment. Factors like macroeconomic news and political statements, often quantified by the CBOE Volatility Index (VIX) or "fear index," significantly impact its price.

Traditional univariate models like ARIMA are ill-equipped for this environment as they cannot incorporate these external variables. Consequently, this project adopts a machine learning approach. As noted by Machine Learning Mastery (2021), ensemble methods like XGBoost can effectively model time-series by reframing the task as a supervised learning problem.

The objective is to develop a short-term XGBoost forecasting model for NVDA. We will test the hypothesis that a model enriched with technical indicators and market sentiment data (VIX) can yield more accurate predictions than models relying on price data alone.

The analysis is divided into three main tasks:

* **Data Acquisition & Cleaning:** Sourcing NVDA (OHLCV) and VIX data; engineering a feature set of technical indicators, lags, and sentiment.

* **Exploratory Data Analysis (EDA):** Visualizing historical trends, volatility, and the NVDA-VIX relationship.

* **XGBoost Modeling:** Implementing, tuning, and evaluating the short-term forecast model.

(Reference: Brownlee, J. (2021). "How to Use XGBoost for Time Series Forecasting." Machine Learning Mastery. Retrieved from https://machinelearningmastery.com/xgboost-for-time-series-forecasting/)

## 2. Data Wrangling and Cleaning

This section covers loading the raw data, cleaning it, and transforming it into a structured format suitable for analysis.

### 2.1. Load Libraries

First, we load all necessary R packages for data processing, analysis, and visualization.

```{r load-libraries, message=FALSE, warning=FALSE}
library(httr)       # For making API calls
library(jsonlite)   # For parsing JSON data
library(dplyr)      # For data manipulation
library(readr)      # For reading/writing CSV files
library(ggplot2)    # For plotting
library(TTR)        # For technical indicators (SMA, RSI, MACD)
library(xgboost)    # For XGBoost modeling
library(zoo)        # For rollapply
```

### 2.2. Data Acquisition (API Call)

We acquire two primary datasets for this analysis.

* **NVIDIA (NVDA) Price Data:** We use httr to call the Alpha Vantage API for NVDA's daily prices. The raw JSON is saved to data/ to avoid repeated API calls.

* **Market Sentiment Data:** We will also load historical data for the CBOE Volatility Index (VIX), which serves as our "fear index."

Both datasets will later be processed and joined by their common date column to create the final feature set for modeling.

```{r data-acquisition, eval=FALSE}
api_key <- "" # Key is removed for privacy
symbol <- "NVDA"
api_url <- paste0("https://www.alphavantage.co/query?function=TIME_SERIES_DAILY",
                  "&symbol=", symbol, "&outputsize=full", "&apikey=", api_key)

response <- GET(api_url)
json_content <- content(response, "text", encoding = "UTF-8")

raw_json_path <- "data/raw_nvda_data.json"
write_lines(json_content, file = raw_json_path)
```

### 2.3. Data Cleaning and Processing

This section involves processing both the NVDA and VIX datasets and then combining them.

First, the raw, nested JSON for NVDA is parsed, converted into a data frame, and cleaned. Columns are renamed for clarity (e.g., X1..open to open), and all values are converted to their proper data types (Date, Numeric).

Second, the downloaded VIX (CBOE Volatility Index) data is loaded from its CSV file. We clean this dataset by renaming its columns, filtering out non-trading days (where the value is . ), and converting its values to Date and Numeric types.

Finally, we perform an inner join on the two datasets using their common date column. This operation aligns both time series and automatically discards any dates where either market was closed (e.g., weekends, holidays). This results in a single, clean, and synchronized dataset ready for analysis.

```{r clean-data, message=FALSE}
# 1. Load and Process NVDA JSON
raw_data <- fromJSON("data/raw_nvda_data.json")
time_series_data <- raw_data$`Time Series (Daily)`

df <- data.frame(
  date = names(time_series_data),
  do.call(rbind, time_series_data),
  row.names = NULL
)

# Rename columns
df_cleaned <- df %>%
  rename(
    open = `X1..open`,
    high = `X2..high`,
    low = `X3..low`,
    close = `X4..close`,
    volume = `X5..volume`
  )

# Convert data types and sort NVDA data
df_nvda <- df_cleaned %>%
  mutate(
    date = as.Date(date),
    open = as.numeric(open),
    high = as.numeric(high),
    low = as.numeric(low),
    close = as.numeric(close),
    volume = as.numeric(volume)
  ) %>%
  arrange(date)

# 2. Load and Process VIX Data
# Data sourced from FRED: https://fred.stlouisfed.org/series/VIXCLS
# Manually downloaded and saved as 'data/VIXCLS.csv'
vix_data <- read_csv("data/VIXCLS.csv")

vix_cleaned <- vix_data %>%
  rename(date = observation_date, vix_close = VIXCLS) %>%
  filter(vix_close != ".") %>% # Filter out non-trading days (marked as ".")
  mutate(
    date = as.Date(date),
    vix_close = as.numeric(vix_close)
  )

# 3. Join Datasets
# An inner join synchronizes the datasets, keeping only dates
# where both the stock market and VIX data are available.
df_final <- inner_join(df_nvda, vix_cleaned, by = "date")

# Save the final joined dataset
write_csv(df_final, "data/processed_joined_data.csv")

# Display summary
cat("Final joined dataset dimensions:", nrow(df_final), "rows and", ncol(df_final), "columns\n")
head(df_final)
```

# 3. Exploratory Data Analysis (EDA)

Before modeling, we explore the combined df_final dataset. The goal is to understand historical trends, volatility patterns, and the key relationships between NVIDIA's stock price and overall market sentiment (VIX).

## 3.1. Basic Summary Statistics

We first examine the summary statistics for the entire joined dataset, which now includes the vix_close column.

```{r summary-stats}
summary(df_final)
```

The summary() output shows our final dataset spans from January 2, 2020 to October 27, 2025. This period, which starts just before the 2020 COVID-19 crash, was selected by the intersection of our available NVDA and VIX data.

* **NVDA Close**: The closing price has ranged from a minimum of $94.31 to a maximum of $1224.40. (Note: NVIDIA executed a 10-for-1 stock split in June 2024, so historical prices reflect pre-split values).

* **VIX Close**: Our market sentiment feature, the VIX, has ranged from a low of $11.86 (indicating market stability) to a high of $82.69. This high value reflects the extreme market panic during the 2020 crash, confirming our dataset includes periods of extreme sentiment.

## 3.2. Handling Stock Split Adjustment

NVIDIA executed a 10-for-1 stock split on June 10, 2024. This creates a structural break in the price data that would distort our time-series analysis and subsequent XGBoost modeling. To ensure continuity, we adjust all pre-split prices by dividing them by 10, making all historical prices comparable to current post-split values.

```{r adjust-split}
# Define the split date
split_date <- as.Date("2024-06-10")

# Adjust pre-split prices
df_adjusted <- df_final %>%
  mutate(
    open = ifelse(date < split_date, open / 10, open),
    high = ifelse(date < split_date, high / 10, high),
    low = ifelse(date < split_date, low / 10, low),
    close = ifelse(date < split_date, close / 10, close)
  )

# Display adjusted summary
cat("Adjusted Summary:\n")
summary(df_adjusted$close)
```

After adjustment, all prices are now on a consistent scale, allowing for accurate feature engineering and modeling.

## 3.3. Historical Price Trend

We visualize the adjusted closing price over time to observe long-term trends and major price movements within our selected 2020-2025 timeframe.

```{r price-trend, fig.width=10, fig.height=5}
ggplot(df_adjusted, aes(x = date, y = close)) +
  geom_line(color = "steelblue") +
  scale_x_date(breaks = seq(as.Date("2020-01-01"), as.Date("2026-01-01"), by = "1 year"),
               date_labels = "%Y") +
  labs(title = "NVIDIA (NVDA) Adjusted Historical Closing Price (2020-2025)",
       x = "Date",
       y = "Adjusted Closing Price (USD)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The plot starkly captures the volatility of the post-2020 era. We can clearly see:

* **The sharp, brief crash during the COVID-19 pandemic in early 2020.**

* **A subsequent explosive growth phase through 2021.**

* **A major correction and bear market throughout 2022.**

* **A rapid recovery and unprecedented bull run from 2023 onward, driven by the generative AI boom, leading to new all-time highs in 2024-2025.**

## 3.4. Volatility Analysis

To better understand the sharp price corrections observed in the previous plot, we calculate the 30-day rolling volatility based on daily returns. This measures risk and price instability over time, helping us identify periods of extreme market stress.

```{r volatility, fig.width=10, fig.height=5}
# Calculate daily returns
df_adjusted <- df_adjusted %>%
  mutate(daily_return = (close - lag(close)) / lag(close))

# Calculate 30-day rolling volatility
df_adjusted <- df_adjusted %>%
  mutate(volatility_30d = zoo::rollapply(daily_return, 
                                         width = 30, 
                                         FUN = sd, 
                                         fill = NA, 
                                         align = "right"))

# Plot volatility
ggplot(df_adjusted, aes(x = date, y = volatility_30d)) +
  geom_line(color = "darkred") +
  scale_x_date(breaks = c(seq(as.Date("2000-01-01"), as.Date("2015-01-01"), by = "2 years"),
                          seq(as.Date("2016-01-01"), as.Date("2026-01-01"), by = "1 year")),
               date_labels = "%Y") +
  labs(title = "NVIDIA (NVDA) 30-Day Rolling Volatility",
       x = "Date",
       y = "Volatility (Standard Deviation of Returns)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The volatility plot clearly confirms the dynamic nature of the stock within our 2020-2025 timeframe. The analysis highlights several key periods of extreme stress:

* **Early 2020:** A major spike in volatility is visible, coinciding with the COVID-19 market crash.

* **Late 2021 - Early 2022:** The most dramatic spike in the entire dataset occurs here. This corresponds perfectly to the severe tech sector correction and bear market of 2022.

* **2023-2025 (AI Boom):** Even during the explosive bull run of the AI boom, the plot shows that volatility did not return to low levels. It remained consistently elevated with frequent, sharp spikes, indicating continuous high risk and investor uncertainty.

This analysis of NVIDIA's inherent volatility strongly supports our decision to include market-wide volatility (the vix_close feature) as a key predictor in our model.

## 3.5. Trading Volume Analysis

Trading volume provides insight into market activity and investor interest, and it is a key feature in our XGBoost model. High volume during price movements can indicate strong conviction behind the trend.

```{r volume-analysis, fig.width=10, fig.height=5}
ggplot(df_adjusted, aes(x = date, y = volume)) +
  geom_line(color = "darkgreen") +
  # Adjust x-axis breaks to match our 2020-2025 data range
  scale_x_date(breaks = seq(as.Date("2020-01-01"), as.Date("2026-01-01"), by = "1 year"),
               date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "NVIDIA (NVDA) Daily Trading Volume (2020-2025)",
       x = "Date",
       y = "Volume") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The plot shows that trading volume remained relatively moderate and stable from 2020 through 2022. A dramatic surge began in 2023-2024, coinciding with the AI boom and NVIDIA's explosive stock price growth.

Peak volumes (confirmed by our summary data to exceed 800 million shares) occurred in 2024, reflecting unprecedented investor interest and market liquidity. This confirms that volume is a highly dynamic feature and justifies its inclusion in our model.

# 3.6. Price vs Market Sentiment(VIX)

```{r price-vix-correlation, fig.width=10, fig.height=7}
library(patchwork) # For combining plots

# Plot 1: NVDA Close Price
p1 <- ggplot(df_adjusted, aes(x = date, y = close)) +
  geom_line(color = "steelblue") +
  labs(title = "NVIDIA Adjusted Close Price", y = "NVDA Price (USD)") +
  theme_minimal()

# Plot 2: VIX Close Price
p2 <- ggplot(df_adjusted, aes(x = date, y = vix_close)) +
  geom_line(color = "darkred") +
  labs(title = "Market Sentiment (VIX Index)", y = "VIX Close") +
  theme_minimal()

# Combine the two plots vertically, aligning the x-axis
p1 + p2 + plot_layout(ncol = 1)
```

The plots reveal a clear inverse correlation, but with an important nuance.

While the single largest VIX spike (the 80+ spike during the COVID-19 crash in early 2020) corresponds to only a moderate price dip for NVIDIA, the biggest NVIDIA price drop (the 2022 bear market) aligns perfectly with the sustained period of high VIX levels (hovering between 20-35) throughout 2022.

Furthermore, the massive bull run for NVIDIA from 2023-2025 corresponds to a period of sustained low VIX, indicating market stability. This visual evidence strongly supports our hypothesis that vix_close is a powerful (and complex) negative predictor for NVDA's price.

# 4. LSTM Modeling
Given the failure of the tabular-based XGBoost model to handle the non-stationary, trended price data, we pivot our approach. We will now replicate the successful strategy from the Python notebook: a Long Short-Term Memory (LSTM) model.

LSTM is a type of Recurrent Neural Network (RNN) specifically designed to learn patterns from sequential data, making it ideal for this problem. We will use the keras package in R.

# 4.1. Setup and Data Preparation
First, we load the keras library and prepare our dataset. We will use only the close price, as in the reference notebook, and split it 80% for training and 20% for testing.

```{r set-data-preparation, fig.width=10, fig.height=7}
library(keras)
library(tensorflow)

# 1. Get the 'close' price data (already split-adjusted)
data <- df_adjusted %>% 
  select(date, close) %>% 
  arrange(date)

# 2. Set the training data length (80% of the data)
training_data_len <- floor(nrow(data) * 0.8)
cat("Total data points:", nrow(data), "\n")
cat("Training points (80%):", training_data_len, "\n")
cat("Test points (20%):", nrow(data) - training_data_len, "\n")
```

# 4.2. Scaling and Sequence Creation
This is the most critical step. We will:

Manually create a MinMaxScaler to scale all data between 0 and 1. We fit this only on the training data.

Define a create_dataset function that takes the scaled data and converts it into 60-day "lookback" windows, just like the Python notebook.

```{r scailing-sequence-creation, fig.width=10, fig.height=7}
# 1. Prepare the scaler based ONLY on training data
train_data_to_scale <- data$close[1:training_data_len]
min_val <- min(train_data_to_scale)
max_val <- max(train_data_to_scale)

cat("Scaler Min:", min_val, "\n")
cat("Scaler Max:", max_val, "\n")

# 2. Define the scaling and inverse-scaling functions
scale_data <- function(x) { (x - min_val) / (max_val - min_val) }
unscale_data <- function(x) { x * (max_val - min_val) + min_val }

# 3. Scale the ENTIRE dataset
scaled_data <- scale_data(data$close)

# 4. Create the 60-day lookback dataset function
create_dataset <- function(dataset, look_back = 60) {
  X <- list()
  y <- list()
  for (i in (look_back + 1):length(dataset)) {
    X[[i - look_back]] <- dataset[(i - look_back):(i - 1)]
    y[[i - look_back]] <- dataset[i]
  }
  list(X = do.call(rbind, X), y = unlist(y))
}

# 5. Create the training dataset
look_back <- 60
train_data_scaled <- scaled_data[1:training_data_len]
train_sequences <- create_dataset(train_data_scaled, look_back)

X_train <- train_sequences$X
y_train <- train_sequences$y

cat("X_train dimensions:", dim(X_train), "\n")
cat("y_train length:", length(y_train), "\n")
```
# 4.3. Build and Train LSTM Model

Now we build the LSTM model using keras. The architecture will match the Python notebook (2 LSTM layers, 2 Dense layers). We must reshape our X_train data into the 3D array that LSTM expects: [samples, timesteps, features].

```{r build-train-lstm, fig.width=10, fig.height=7}
# 1. Reshape X_train for LSTM [samples, timesteps, features]
X_train_reshaped <- array_reshape(X_train, c(dim(X_train)[1], dim(X_train)[2], 1))
cat("Reshaped X_train dimensions:", dim(X_train_reshaped), "\n")

# 2. Build the Keras model
model <- keras_model_sequential()

model %>% 
  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(look_back, 1))

model %>% 
  layer_lstm(units = 50, return_sequences = FALSE)

model %>% 
  layer_dense(units = 25)

model %>% 
  layer_dense(units = 1)

# 3. Compile the model
model %>% compile(
  optimizer = 'adam',
  loss = 'mean_squared_error'
)

summary(model)

# 4. Train the model
cat("Starting model training...\n")
history <- model %>% fit(
  x = X_train_reshaped,
  y = y_train,
  batch_size = 32,
  epochs = 1,
  verbose = 1
)
```