---
title: "Lab1 Jooho Kim"
ASUID: "1234806471"
course: "DAT301-74297"
output: pdf_document
date: "2025-09-21"
---

```{r load_data}
df = read.csv("Absenteeism_at_work.csv", sep=";", header=TRUE)
head(df)
```
```{r scatter_height_weight}
plot(df$Weight, df$Height,
     xlab = "Weight",
     ylab = "Height",
     main = "Scatter Plot of Height vs Weight")
```
There does not appear to be a strong linear relationship between height and weight.

```{r hist_hours_absence}
hist(df$Absenteeism.time.in.hours,
     xlab = "Hours of Absence",
     main = "Histogram of Hours of Absence")
```

The histogram shows the distribution of absence hours, with most values clustered at lower hours.

```{r hist_age}
hist(df$Age,
     xlab = "Age",
     main = "Histogram of Age")
```

The histogram shows the age distribution of employees with recorded absences.

```{r bar_month_absence}
total_hours_by_month <- tapply(df$Absenteeism.time.in.hours,
                               df$Month.of.absence,
                               sum)

barplot(total_hours_by_month,
        xlab = "Month",
        ylab = "Total Hours of Absence",
        main = "Total Absence Hours by Month")
```

The bar plot shows how total absence hours vary across months.

```{r box_smoker}
boxplot(df$Absenteeism.time.in.hours ~ df$Social.smoker,
        xlab = "Social Smoker",
        ylab = "Hours of Absence",
        main = "Absence Hours by Social Smoker",
        col = c("lightblue", "lightgreen"),
        names = c("Non-Smoker", "Smoker"))

legend("topright",
       legend = c("Non-Smoker", "Smoker"),
       fill = c("lightblue", "lightgreen"))
```

The box plot compares absence hours between social smokers and non-smokers.

```{r box_drinker}
boxplot(df$Absenteeism.time.in.hours ~ df$Social.drinker,
        xlab = "Social Drinker",
        ylab = "Hours of Absence",
        main = "Absence Hours by Social Drinker",
        col = c("orange", "lightblue"),
        names = c("Non-Drinker", "Drinker"))

legend("topright",
       legend = c("Non-Drinker", "Drinker"),
       fill = c("orange", "lightblue"))
```

The box plot compares absence hours between social drinkers and non-drinkers.

